---
title: "Predictive modeling of customer bookings"
author: "Rounak Saha"
date: "2025-01-26"
output:
  pdf_document: default
  html_document: default
---

# Introduction
Customers are more empowered than ever because they have access to a wealth of information at their fingertips. This is one of the reasons the buying cycle is very different to what it used to be. Today, if you’re hoping that a customer purchases your flights or holidays as they come into the airport, you’ve already lost! Being reactive in this situation is not ideal; airlines must be proactive in order to acquire customers before they embark on their holiday.

### Task
With your predictive model, it is important to interpret the results in order to understand how “predictive” the data really was and whether we can feasibly use it to predict the target outcome (customers buying holidays). Therefore, you should evaluate the model's performance and output how each variable contributes to the predictive model's power.

```{r, echo = TRUE, eval = TRUE, results="hide", fig.keep = "none"}
# Install all the neccessary packages 
options(repos = c(CRAN = "https://cloud.r-project.org"))
install.packages("tidyverse")
install.packages("caret")        # split the data
install.packages("caTools")      # K-fold cross validation
install.packages("xgboost")      # Xgboost ml model
install.packages("countrycode")  #Convert Country Names and Country Codes
install.packages("ggplot2")      # plot the graph
install.packages("gridExtra") #arrange multiple grid-based plots on a page
install.packages("e1071") #for log transformation
install.packages("fastDummies") # for one-hot encoding
```

```{r, echo = TRUE, eval = TRUE, results="hide",message=FALSE, fig.keep = "none", warning=FALSE}
# Load the library 
library(tidyverse)
library(caret)
library(caTools)
library(xgboost)
library(countrycode)
library(ggplot2)
library(gridExtra)
library(e1071)
library(fastDummies)
```

About the data :
To provide more context, below is a more detailed data description, explaining exactly what each column means:

num_passengers = number of passengers travelling

sales_channel = sales channel booking was made on

trip_type = trip Type (Round Trip, One Way, Circle Trip)

purchase_lead = number of days between travel date and booking date

length_of_stay = number of days spent at destination

flight_hour = hour of flight departure

flight_day = day of week of flight departure

route = origin = destination flight route

booking_origin = country from where booking was made

wants_extra_baggage = if the customer wanted extra baggage in the booking

wants_preferred_seat = if the customer wanted a preferred seat in the booking

wants_in_flight_meals = if the customer wanted in-flight meals in the booking

flight_duration = total duration of flight (in hours)

booking_complete = flag indicating if the customer completed the booking

```{r}
#Load the dataset and explore 
customer_booking <- read_csv("customer_booking.csv")
#View(customer_booking)

head(customer_booking)
str(customer_booking)

unique(customer_booking$sales_channel)
unique(customer_booking$trip_type)
unique(customer_booking$route)
unique(customer_booking$booking_origin)
```

# Data Cleaning and Manupulation 

```{r, warning=FALSE}
# Encoding the categorical variables
#Label Encoding
customer_booking$sales_channel <- as.numeric(factor(customer_booking$sales_channel, 
                                                    levels = unique(customer_booking$sales_channel)))-1
#,
                                            #levels = c('Internet', 'Mobile'),
                                            #labels = c(1,2))
#Mapping Encoding
trip_type_mapping <- c("RoundTrip" = 0, "OneWay" = 1, "CircleTrip" = 2)
customer_booking$trip_type <- trip_type_mapping[customer_booking$trip_type]
customer_booking$trip_type<- as.numeric(customer_booking$trip_type)

# Replace specific country name
customer_booking$booking_origin[customer_booking$booking_origin == "Myanmar (Burma)"] <- "Myanmar"

# The booking origin column also has many unique values, 
# but because I don't want to delete the information on the origin of the booking, 
# I will change the value of the booking origin, which initially contains 
# the name of the country to the name of the continent

customer_booking$booking_origin <- countrycode(customer_booking$booking_origin, "country.name", "continent")
customer_booking$booking_origin[is.na(customer_booking$booking_origin)] <- "Others"

#customer_booking$booking_origin <- factor(customer_booking$booking_origin, 
                                        #  levels = c("Oceania","Asia","Europe","Americas", "Africa","Others"),
                                         # labels = c(1:6))
#(its better to do one-hot encoding when there are nominal cat values in col and more in number)

# Drop columns that has many unique values
length(unique(customer_booking$route))
customer_booking <- customer_booking[,-8]

#Make a new feature Because we want to know customer behavior to have a trip on holiday (weekend), 
#so let's make a feature called is_weekend. if the flight day is Saturday or Sunday we give is_weekend value = 1, 
#for another flight day we give it 0
customer_booking <- customer_booking %>%
  mutate(is_weekend = ifelse(flight_day %in% c("Sat", "Sun"), 1, 0))
customer_booking <- customer_booking[,-7]

summary(customer_booking)


# converting variables to factors
#customer_booking$wants_extra_baggage <- as.factor(customer_booking$wants_extra_baggage)
#customer_booking$wants_preferred_seat <- as.factor(customer_booking$wants_preferred_seat)
#customer_booking$wants_in_flight_meals <- as.factor(customer_booking$wants_in_flight_meals)
#customer_booking$is_weekend <- as.factor(customer_booking$is_weekend)
#customer_booking$booking_complete<- as.factor(customer_booking$booking_complete)

str(customer_booking)
```
# Data Transformation and Visualization 
```{r, warning=FALSE}
str(customer_booking)

#check the distribution on numerical data
a1<- ggplot(customer_booking,                       # Initializes ggplot() using the dataset customer_booking
            aes(x = num_passengers)) +              #Specifies that num_passengers is the variable for the x-axis
  geom_histogram(aes(y = ..density..),              #By default, histograms show counts (frequency of occurrences).
                                                    #..density.. ensures that the histogram is scaled to show density instead of raw counts, 
                                                    # making it comparable with a density curve
                 bins = 30,                         #Divides the x-axis range into 30 bins (intervals) for better granularity.
                 fill = "lightblue",                #Fills the bars with a light blue color.
                 color = "black") +                 #Adds black borders to each bin for better visibility.
  geom_density(                                     #Adds a density curve (smoothed probability distribution). 
    color = "red",                                  #Makes the curve red for distinction.
    size = 1.2)                                     #Increases the line thickness for better visibility.


a2<- ggplot(customer_booking, aes(x = purchase_lead)) +
  geom_histogram(aes(y = ..density..), 
                 bins = 30, 
                 fill = "lightblue", 
                 color = "black") +
  geom_density(color = "red", size = 1.2)

a3<- ggplot(customer_booking, aes(x = length_of_stay)) +
  geom_histogram(aes(y = ..density..), 
                 bins = 30, 
                 fill = "lightblue", 
                 color = "black") +
  geom_density(color = "red", size = 1.2)

a4<- ggplot(customer_booking, aes(x = flight_hour)) +
  geom_histogram(aes(y = ..density..), 
                 bins = 30, 
                 fill = "lightblue", 
                 color = "black") +
  geom_density(color = "red", size = 1.2)

a5<- ggplot(customer_booking, aes(x = flight_duration)) +
  geom_histogram(aes(y = ..density..), 
                 bins = 30, 
                 fill = "lightblue", 
                 color = "black") +
  geom_density(color = "red", size = 1.2)

grid.arrange(a1,a2,a3,a4,a5, nrow= 3, ncol=2)

###### OR #######

#variables1 <- c("num_passengers", "purchase_lead", 
#               "length_of_stay", "flight_hour", "flight_duration")

#plots <- lapply(variables1, function(var){
#  ggplot(customer_booking, aes_string(x= var)) +
#    geom_histogram(aes(y = ..density..), 
#                   bins = 30,
#                   fill= "lightblue",
#                   color = "black") +
#    geom_density(color= "red", size= 1.2)+
#    ggtitle(var)
# })
# grid.arrange(grobs= plots, nrow=3, ncol=2)
```

```{r}
# Copy original data
df_transformed <- customer_booking

# Checking the skewness of the numeric variables
skewness(df_transformed$num_passengers)
skewness(df_transformed$purchase_lead)
skewness(df_transformed$length_of_stay)
skewness(df_transformed$flight_hour)
skewness(df_transformed$flight_duration)

#skewness(df_transformed$num_passengers) #log
#[1] 2.690747
#skewness(df_transformed$purchase_lead) #sqrt
#[1] 1.652936
#skewness(df_transformed$length_of_stay)# log
#[1] 5.274426
#skewness(df_transformed$flight_hour). #no need
#[1] 0.3965994
#skewness(df_transformed$flight_duration)
#[1] -0.3600581

#Skewness Range	Recommended Transformation
#0 to ±0.5	Already normal (No transformation needed)
#0.5 to ±1.5	Log Transformation (log1p(x))
#1.5 to ±3.0	Square Root Transformation (sqrt(x))
#Above 3.0	Box-Cox or Log Transformation (log1p(x))
  
# Apply Log Transformation to Selected Columns
cols1 <- c("num_passengers", "length_of_stay")

for (col in cols1) {
  df_transformed[[col]] <- log1p(df_transformed[[col]])  # log(x + 1) to avoid log(0)
}                                            #[[col]] allows column selection dynamically inside a loop.

# check skewness 
skewness(df_transformed$num_passengers)
skewness(df_transformed$length_of_stay)

# Apply Square Root Transformation to Selected Columns
df_transformed$purchase_lead <- sqrt(df_transformed$purchase_lead)
skewness(df_transformed$purchase_lead)

# there is no need to transform the flight hour column

# Exponential Transformation to flight_duration column since the data is left skewed
df_transformed$flight_duration <- exp(df_transformed$flight_duration)
summary(df_transformed$flight_duration)
# plot boxplot 
# Find outliers 
# Let's see outlier on numeric column
variables2 <- c("num_passengers", "purchase_lead", 
                "length_of_stay", "flight_hour", "flight_duration")

plots2 <- lapply(variables2, function(var){
  ggplot(df_transformed, aes_string(x= var)) +
    geom_boxplot()
})
grid.arrange(grobs = plots2, nrow= 3, ncol= 2)

# as we can see, on purchase lead and length of stay have a lot of outlier values, 
# so we will delete outlier wtih zscore so that not many columns are wasted
cols2 <- c("num_passengers", "purchase_lead", 
           "length_of_stay", "flight_hour", "flight_duration")

for(col in cols2){
  zscore <- abs(scale(df_transformed[[col]]))
  df_transformed <- df_transformed[zscore <3,]
}

#Now checking the boxplot again
variables3 <- c("num_passengers", "purchase_lead", 
                "length_of_stay", "flight_hour", "flight_duration")

plots3 <- lapply(variables3, function(var){
  ggplot(df_transformed, aes_string(x= var)) +
    geom_boxplot()
})
grid.arrange(grobs = plots3, nrow= 3, ncol= 2)

# One-hot encoding for booking_origin column
df_encoded <- dummy_cols(df_transformed, select_columns = "booking_origin", 
                         remove_selected_columns = TRUE)
df_encoded
```

# Make a machine learning Model- XGboost
```{r}
#Split Data
str(df_encoded)

# Convert all integer columns to numeric
df_encoded[] <- lapply(df_encoded, as.numeric) ##[] Preserves the data frame structure while applying lapply().
                                               #Ensures that all columns are converted to numeric without changing df_encoded to a list.

#Splitting the data
set.seed(1234)
split <- sample.split(df_encoded$booking_complete, SplitRatio = 0.8)
train_set <- subset(df_encoded, split == TRUE)
test_set <- subset(df_encoded, split ==FALSE)

#Fitting Xgboost to training set
classifier <- xgboost(data = as.matrix(train_set[-11]), 
                      label = train_set$booking_complete, nrounds = 10)

# Predicting the Test set results
y_pred <- predict(classifier, newdata = as.matrix(test_set[-11]))
y_pred <- (y_pred >= 0.4)

# Making the Confusion Matrix
cm <- table(test_set$booking_complete, y_pred)
Acc <- (cm[1,1]+cm[2,2])/ sum(cm)

# Applying k-Fold Cross Validation
set.seed(123)
folds = createFolds(df_encoded$booking_complete, k = 5)
cv = lapply(folds, function(x) {
  training_fold = df_encoded[-x, ]
  test_fold = df_encoded[x, ]
  classifier = xgboost(data = as.matrix(training_fold[-11]), 
                       label = training_fold$booking_complete, nrounds = 10)
  y_pred = predict(classifier, newdata = as.matrix(test_fold[-11]))
  y_pred = (y_pred >= 0.4)
  cm = table(test_fold$booking_complete, y_pred)
  accuracy = (cm[1,1] + cm[2,2]) / (cm[1,1] + cm[2,2] + cm[1,2] + cm[2,1])
  recall = cm[2,2] / (cm[2,1] + cm[2,2])
  return(c(accuracy, recall))
})

cv_results_df = do.call(rbind, cv)
colnames(cv_results_df) = c("Accuracy", "Recall")
                            
mean_Accuracy = mean(cv_results_df[, "Accuracy"])
mean_Accuracy
mean_recall = mean(cv_results_df[, "Recall"])
mean_recall
```
# Feature Importance 
```{r}
# Get feature importance
feature_importance <- xgb.importance(model = classifier)

# Convert to dataframe and sort by importance
feature_importance_df <- feature_importance %>%
  arrange(desc(Gain)) %>%  # 'Gain' is the most important metric
  head(10)  # Select top 10 features

# Plot feature importance
ggplot(feature_importance_df, aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +  # Horizontal bar plot
  labs(title = "Feature Importance (Top 10)", x = "Features", y = "Importance (Gain)") +
  theme_minimal()

```




